Lichat, A Lightweight Chat Protocol

### 2022-01-23 els2022-lichat/paper.tex
Minor section

\section{Introduction}\label{introduction}
The IRC (Internet Relay Chat) protocol has long stood the test of time as a reliable protocol for text chat communication. Large networks and a plethora of mature clients still exist and are in use today. However, The IRC protocol is also plagued by a number of issues that stem from its historical design and largely distributed evolution. In particular, the protocol has many different, incompatible extensions, and features many idiosyncrasies that make it difficult to write new clients or servers.

@@

Over time many other chat systems such as AIM, ICQ, Skype, have come and largely gone, though all of these focus on a different model of operation than IRC. They focus on a friends-list based approach where people talk to each other directly, rather than relying on sometimes huge public chat rooms.

@@

\begin{listing}[h]
\begin{minted}[fontsize=\scriptsize]{common-lisp}
> (connect :id 0 :clock 5243 :from "tester" :version "1.0" :extensions ())
< (connect :id 0 :clock 5243 :from "tester" :version "1.0" :extensions ())
> (join :id 0 :clock 5243 :from "tester" :channel "lichatters")
< (join :id 0 :clock 5243 :from "tester" :channel "lichatters")
< (message :id 0 :clock 5244 :from "shinmera" :channel "lichatters" :text "Welcome!")
> (message :id 0 :clock 5246 :from "tester" :channel "lichatters" :text "Hiya")
< (message :id 0 :clock 5246 :from "tester" :channel "lichatters" :text "Hiya")
\end{minted}
\caption{A basic protocol exchange.}
\label{lst:basic-exchange}
\end{listing}

@@

The permissions system in general is very simple: each channel holds a map where each update type is associated with with either a list of permitted users, or a list of denied users. As all communication between users must occur through channels, this provides a simple, but expressive model to handle moderation. For instance, banning a user simply requires removing their ability to send a \code{join} message to the channel.

@@

Further, we required the use of a null byte as an end of message marker. This is useful so that a server or client can restore the stream of messages should a parser failure occur, or should the update be rejected due to excessive length.

@@

>>>>

\section{Introduction}\label{introduction}
The IRC (Internet Relay Chat) protocol has long stood the test of time as a reliable protocol for text chat communication. Large networks and a plethora of mature clients still exist and are in use today. However, the IRC protocol is also plagued by a number of issues that stem from its historical design and largely distributed evolution. In particular, the protocol has many different, incompatible extensions, and features many idiosyncrasies that make it difficult to write new clients or servers.

@@

Over time many other chat systems such as AIM, ICQ, and Skype have come and largely gone, though all of these focus on a different model of operation than IRC. They focus on a friends-list based approach where people talk to each other directly, rather than relying on sometimes huge public chat rooms.

@@

\begin{listing}[h]
\begin{minted}[fontsize=\scriptsize]{common-lisp}
> (connect :id 0 :clock 5243 :from "tester" :version "1.0" :extensions ())
< (connect :id 0 :clock 5243 :from "tester" :version "1.0" :extensions ())
> (join :id 1 :clock 5243 :from "tester" :channel "lichatters")
< (join :id 1 :clock 5243 :from "tester" :channel "lichatters")
< (message :id 37 :clock 5244 :from "shinmera" :channel "lichatters" :text "Welcome!")
> (message :id 2 :clock 5246 :from "tester" :channel "lichatters" :text "Hiya")
< (message :id 2 :clock 5246 :from "tester" :channel "lichatters" :text "Hiya")
\end{minted}
\caption{A basic protocol exchange.}
\label{lst:basic-exchange}
\end{listing}

@@

The permissions system in general is very simple: each channel holds a map where each update type is associated with either a list of permitted users, or a list of denied users. As all communication between users must occur through channels, this provides a simple, but expressive model to handle moderation. For instance, banning a user simply requires removing their ability to send a \code{join} message to the channel.

@@

Further, we required the use of a null byte as an end of message marker. This is useful so that a server or client can restore the stream of messages should a parser failure occur, or should the update be rejected due to excessive length. We also specified that an optional, missing field is equivalent to specifying the field to be \code{nil}, and vice-versa. This simplifies initialising and dealing with optional values a lot.

@@

Aside from the wire protocol and message type hierarchy, each client follows a pretty simple scheme as outlined in \autoref{lst:client}. Particularly simplifying is the fact that protocol-level pings are sent by the server, removing the need for complex timeout tracking in the client.

\begin{listing}
\begin{minted}[fontsize=\footnotesize]{common-lisp}
(defun connect ()
  (socket-connect)
  (send 'connect :from username)
  (let ((reply (make-update (read-list))))
    (etypecase reply
      (connect
       (setf username (from reply))
       (setf supported-extensions (extensions reply))))))

(defun skip ()
  (loop until (= 0 (read-byte)))
  (continue))

(defun reconnect ()
  (socket-disconnect)
  (connect)
  (continue))

(defun message-loop ()
  (loop (with-simple-restart (continue)
          (handle (make-update (read-list))))))

(defun main ()
  (connect)
  (handler-bind ((syntax-error #'skip)
                 (update-malformed #'continue)
                 (timeout #'reconnect))
    (message-loop)))
\end{minted}
\caption{A pseudo-code example of a client's main loop}
\label{lst:client}
\end{listing}
